---
title: "Making sense of Paramecium plasticity and growth across temperatures"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd('C:/Users/kylec/Documents/GitHub/FR_n-prey-at-a-time/ms/PlasticityandGrowth_Paramecium/GAM_Analysis/')

library(dplyr); library(ggplot2); library(cowplot); library(mgcv); library(gratia); library(ggbiplot); library(vegan)




```

## Introduction

One goal of our project is to gain a better understanding of both the morphological and growth plasticity of paramecia in response to temperature. One reason for this is that the changes we see in paramecia under our warm and cold treatments are likely to be attributable partly to plasticity. Another reason is that plasticity is likely to play an important role in determining which outcrossed lines in our original evolution populations were likely to do well at high and low and temperatures. For example, some outcrossed lines may perform better than others at higher versus colder temperatures and vice versa. Moreover, there are some questions that we might be able to examine about plasticity in morphology and growth that are broadly interesting. For example, is the degree of plasticity in morphology related to the degree of plasticity in growth across temperatures? What changes in morphological traits are associated with changes in growth in paramecium?

## What did we do?

The data we have to attempt to answer these questions comes from two experiments that we performed last year. The first was our TPC experiment for the 20 outcrossed lines that we used to start our original evolution populations. The second was a plasticity experiment in which we acclimated each of the 20 lines across different temperatures and then video phenotyped them to examine morphological and phenotypic plasticity across temperatures.

## Characterizing Morphological and Behavioral Plasticity

The first thing we can do is try to characterize the morphological and behavioral plasticity of the paramecia aross temperatures. First, we will load the data and make some modifications to add some necessary information that is currently in the video file names. We will also drop the measurements from the highest temperature since very few cells survived at the highest temperature over the acclimation period.

```{r}

morph_data <- read.csv('Plast_StartPop_Data.csv')

morph_data <- morph_data %>% mutate(Line = as.factor(sapply(strsplit(morph_data$file, split = "_"),'[[', 4)),
                                      Temperature = as.numeric(sapply(strsplit(morph_data$file, split = "_"),'[[', 3)),
                                    LineTemp = paste(Line, Temperature, sep = "_"))

morph_data <- filter(morph_data, Temperature < 37)

```

Now we can look at various morphological and behavioral traits and characterize their plasticity with temperature (we will focus on length, width, aspect ratio, and speed). To do so, I will use Generalized Additive Models (GAMs) as these models are capable of fitting flexible, nonlinear relationships when the forms of these relationships are not necessarily known *a priori*. 

### Length

First, we can look at a plot of paramecium length across temperatures.

```{r}

ggplot(data = morph_data, aes(x = Temperature, y = mean_major)) + geom_point() + geom_smooth(method = 'gam', formula = y ~ s(x, k = 5, bs = 'tp'), se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5)

```
There is a lot going on, but, what we see generally is a tendency towards smaller lengths (mean major axis) as temperature increases with a slight increase in length at the highest temperature. Now we can run the formal GAMs and look at the resulting summary and plots.

```{r}

gam_length <- gam(formula = mean_major ~ Line + s(Temperature, by = Line, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

summary(gam_length)

for (i in 1:length(smooths(gam_length))) {
  out <- draw(gam_length, select = i, residuals = TRUE)     
  print(out) }

```

In general, these fits look pretty good. We can also examine whether including the interaction between line and temperature (i.e. a GxE interaction) improves the fit of the model to the data. We will fit a model without the interaction and then compare the AIC values of the two models.

```{r}

gam_length_noint <- gam(formula = mean_major ~ Line + s(Temperature, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

AIC(gam_length, gam_length_noint)

```

Overall, we see that the GAM including the interaction has a much, much lower AIC score telling us that including the interaction term increases the predictive ability of the model substantially ($\Delta$AIC = `r AIC(gam_length_noint) - AIC(gam_length)`).  

### Width

Next, we can look at paramecium width across temperatures.

```{r}

ggplot(data = morph_data, aes(x = Temperature, y = mean_minor)) + geom_point() + geom_smooth(method = 'gam', formula = y ~ s(x, k = 5, bs = 'tp'), se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5)

```

In general, we see a similar trend as the one we saw with length -- paramecia are generally getting less wide as temperatures increase. Again, we will run GAMs and look at the results. 

```{r}

gam_width <- gam(formula = mean_minor ~ Line + s(Temperature, by = Line, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

summary(gam_width)

for (i in 1:length(smooths(gam_width))) {
  out <- draw(gam_width, select = i, residuals = TRUE)     
  print(out) }

```

Again, we will assess whether the inclusion of the interaction terms improves the model's ability to predict the data.

```{r}

gam_width_noint <- gam(formula = mean_minor ~ Line + s(Temperature, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

AIC(gam_width, gam_width_noint)

```
And, again we see that AIC score is much lower for the model including the interaction ($\Delta$AIC = `r AIC(gam_width_noint) - AIC(gam_width)`).

### Aspect Ratio

Now we will examine how the aspect ratio of the paramecium changes with temperature. 

```{r}

ggplot(data = morph_data, aes(x = Temperature, y = mean_ar)) + geom_point() + geom_smooth(method = 'gam', formula = y ~ s(x, k = 5, bs = 'tp'), se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5)

```

Overall, across the different lines, the general trend is for an overall increase in aspect ratio with temperature. Fitting a GAM gives:

```{r}

gam_ar <- gam(formula = mean_ar ~ Line + s(Temperature, by = Line, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

summary(gam_ar)

for (i in 1:length(smooths(gam_ar))) {
  out <- draw(gam_ar, select = i, residuals = TRUE)     
  print(out) }

```

Again, we will assess whether the inclusion of the interaction terms improves the model's ability to predict the data.

```{r}

gam_ar_noint <- gam(formula = mean_ar ~ Line + s(Temperature, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

AIC(gam_ar, gam_ar_noint)

```

And, again we see that AIC score is much lower for the model including the interaction ($\Delta$AIC = `r AIC(gam_ar_noint) - AIC(gam_ar)`).

### Speed

For our final phenotype, we will look at speed.

```{r}

ggplot(data = morph_data, aes(x = Temperature, y = gross_speed)) + geom_point() + geom_smooth(method = 'gam', formula = y ~ s(x, k = 5, bs = 'tp'), se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5)

```

As for aspect ratio, we some general trends towards faster paramecium at higher temperatures with some exceptions. Again, we can fit a GAM with an interaction between temperature and line.

```{r}

gam_speed <- gam(formula = gross_speed ~ Line + s(Temperature, by = Line, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

summary(gam_speed)

for (i in 1:length(smooths(gam_speed))) {
  out <- draw(gam_speed, select = i, residuals = TRUE)     
  print(out) }

```

Again, we will assess whether the inclusion of the interaction terms improves the model's ability to predict the data.

```{r}

gam_speed_noint <- gam(formula = gross_speed ~ Line + s(Temperature, k = 5, bs = 'tp'), data = morph_data, method = 'REML')

AIC(gam_speed, gam_speed_noint)

```

And, again we see that AIC score is much lower for the model including the interaction ($\Delta$AIC = `r AIC(gam_speed_noint) - AIC(gam_speed)`).

### Examining the structure of morphological plasticity in Paramecium

From our previous results, we know that genetic variation in the paramecia is present for size, aspect ratio, and speed. We also know that variation in length and width are correlated and aspect ratio and speed, but these are not correlated with one another. One question we might be interested in is whether plasticity follows a similar pattern? To examine this, we can look at some plots for each outcrossed line that assess the relationship within each line between the morphological and movement variables and temperature.

First, we can look at length and width.

```{r}

ggplot(data = morph_data, aes(x = mean_major, y = mean_minor)) + geom_point(aes(color = Temperature)) + geom_smooth(method = 'lm', se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5) + 
  scale_color_viridis_b()

```

Across lines, it does appear to be the case that shorter cells are also less wide. Furthermore, looking at the relationships with temperature, many of the lines tend to have the longest and widest cells at low temperatures and the shortest, thinnest cells at high temperatures (e.g. G69). There also is some evidence here that the degree of plasticity differs across the outcrossed lines. For example, lines such as G69 and G16 span a large range of the x-axis across temperatures, whereas, lines such as G77 and G44 occupy a shorter range along the x-axis. 

Now let's look at length and its relationship with aspect ratio. 

```{r}

ggplot(data = morph_data, aes(x = mean_major, y = mean_ar)) + geom_point(aes(color = Temperature)) + geom_smooth(method = 'lm', se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5) + 
  scale_color_viridis_b()

```

Here, we see generally shallower slopes in the relationship between length and aspect ratio.

Let's also look at length and speed.

```{r}

ggplot(data = morph_data, aes(x = mean_major, y = gross_speed)) + geom_point(aes(color = Temperature)) + geom_smooth(method = 'lm', se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5) + 
  scale_color_viridis_b()

```

Here, we do see that some outcrossed lines seem to show a relationship between speed and length, but, again, the relationships aren't quite as consistent as the relationships between length and width.

We can also look at the relationships between width, aspect ratio, and speed.

```{r}

ggplot(data = morph_data, aes(x = mean_minor, y = mean_ar)) + geom_point(aes(color = Temperature)) + geom_smooth(method = 'lm', se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5) + 
  scale_color_viridis_b()

```

```{r}

ggplot(data = morph_data, aes(x = mean_minor, y = gross_speed)) + geom_point(aes(color = Temperature)) + geom_smooth(method = 'lm', se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5) + 
  scale_color_viridis_b()

```

In the width and aspect ratio plots we see a weak negative relationship between width and aspect ratio as we might expect. This appears to translate to negative relationships between width and speed in some outcrossed lines but not others.

Last, we can look at the relationship between aspect ratio and speed.

```{r}

ggplot(data = morph_data, aes(x = mean_ar, y = gross_speed)) + geom_point(aes(color = Temperature)) + geom_smooth(method = 'lm', se = FALSE) + 
  facet_wrap(facets = "Line", nrow = 4, ncol = 5) + 
  scale_color_viridis_b()

```

Here, again, we see fairly strong and consistent relationships between aspect ratio and speed as we had found previously across outcrossed lines in length and width. There is also a tendency for the highest aspect ratios and speeds for each outcrossed line to be at the higher temperatures. 

What does this mean overall? I think that it supports the idea that plasticity is occurring on the same general axes within outcrossed lines as we see for genetic variation across outcrossed lines. That is, there are strong correlations in plasticity between length and width and between aspect ratio and speed, but weaker correlations across the combinations of the variables just as we saw with variation across outcrossed lines at a single temperature.

## Quantifying Plasticity

Now that we have looked at how the different outcrossed lines are responding to temperature in their morphology and movement, we want to look at how plastic each of the outcrossed lines are in their phenotypes. To do this, we will first define the degree of plasticity as the range of the mean trait across the temperature range for each outcrossed line. This feels like a fairly intuitive, ad hoc way to define the degree of plasticity.

```{r}

### get data for length

summ_length <- morph_data %>% group_by(Line, Temperature) %>% summarise(length = median(mean_major)) %>% ungroup() %>% group_by(Line) %>% 
  summarise(mean_length = median(length),
            max_length = max(length),
            min_length = min(length),
            range_length = max_length - min_length,
            stand_range_length = range_length/max_length)

### predictions for width

summ_width <- morph_data %>% group_by(Line, Temperature) %>% summarise(width = median(mean_minor)) %>% ungroup() %>% group_by(Line) %>% 
  summarise(mean_width = mean(width),
            max_width = max(width),
            min_width = min(width),
            range_width = max_width - min_width,
            stand_range_width = range_width/max_width)

### predictions for aspect ratio

summ_ar <- morph_data %>% group_by(Line, Temperature) %>% summarise(ar = median(mean_ar)) %>% ungroup() %>% group_by(Line) %>% 
  summarise(mean_ar = mean(ar),
            max_ar = max(ar),
            min_ar = min(ar),
            range_ar = max_ar - min_ar,
            stand_range_ar = range_ar/max_ar)

### predictions for speed

summ_speed <- morph_data %>% group_by(Line, Temperature) %>% summarise(speed = median(gross_speed)) %>% ungroup() %>% group_by(Line) %>% 
  summarise(mean_speed = mean(speed),
            max_speed = max(speed),
            min_speed = min(speed),
            range_speed = max_speed - min_speed,
            stand_range_speed = range_speed/max_speed)

### put these together into a single data frame

plast_morph_data <- full_join(summ_length, summ_width, by = "Line")

plast_morph_data <- full_join(plast_morph_data, summ_ar, by = "Line")

plast_morph_data <- full_join(plast_morph_data, summ_speed, by = "Line")


```

Now that we have quantified plasticity in each of the phenotypes, we can ask whether plasticity in certain phenotypes are related to one another. That is, are outcrossed lines that are more plastic in length also more plastic in width, aspect ratio, or speed?

```{r}

### length versus width plasticity

ggplot(data = plast_morph_data, aes(x = range_length, y = range_width)) + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + ggtitle('Length v. Width Plasticity')

summary(lm(range_width ~ range_length, data = plast_morph_data))

### length versus ar plasticity 

ggplot(data = plast_morph_data, aes(x = range_length, y = range_ar)) + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + ggtitle('Length v. Aspect Ratio Plasticity')

summary(lm(range_ar ~ range_length, data = plast_morph_data))

### length versus speed plasticity 

ggplot(data = plast_morph_data, aes(x = range_length, y = range_speed)) + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + ggtitle('Length v. Speed Plasticity')

summary(lm(range_speed ~ range_length, data = plast_morph_data))

### width versus aspect ratio plasticity

ggplot(data = plast_morph_data, aes(x = range_width, y = range_ar)) + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + ggtitle('Width v. Aspect Ratio Plasticity')

summary(lm(range_ar ~ range_width, data = plast_morph_data))

### width versus speed plasticity

ggplot(data = plast_morph_data, aes(x = range_width, y = range_speed)) + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + ggtitle('Width v. Speed Plasticity')

summary(lm(range_speed ~ range_width, data = plast_morph_data))


### aspect ratio versus speed plasticity

ggplot(data= plast_morph_data, aes(x = range_ar, y = range_speed)) + geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) + ggtitle('Aspect Ratio v. Speed Plasticity')

summary(lm(range_speed ~ range_ar, data = plast_morph_data))


``` 

These regressions show fairly strong evidence in our data that plasticity in length and width are associated with one another as well as plasticity in aspect ratio and speed, but not among any of the variables between these groups. This suggests that outcrossed lines can vary plastically in size while also varying plastically in aspect ratio/speed either a lot or a little. Again, since we know that these axes of variation show some amount of genetic independence, this suggests that the plasticity in these two sets of traits also varies independently. 

## Growth across temperatures

Now, we want to determine whether there are relationships between the phenotypes of outcrossed lines across temperatures and the population growth rates of the outcrossed lines. First, we will load the growth rate data and make some modifications and then combine the growth rate and phenotype data. To combine the data, we will use the means of the phenotypes as the 'observed' phenotype of the outcrossed line at each temperature.

```{r}

growth_data <- read.csv('StartPop_TPC.csv')

growth_data <- growth_data %>% filter(!is.na(Growth.Rate.Hours) & !is.infinite(Growth.Rate.Hours))

growth_data <- growth_data %>% filter(Genotype != 'blank')

growth_data$Genotype <- as.factor(paste0('G', growth_data$Genotype))

colnames(growth_data)[which(colnames(growth_data) == 'Genotype')] <- 'Line'

### take a look at the growth rate data 

ggplot(data = growth_data, aes(x = Temperature, y = Growth.Rate.Hours, color = Line)) + 
  geom_point() + geom_smooth(method = 'gam', formula = y ~ s(x, k = 6, bs = 'tp'), se = FALSE)

```

```{r}

### set up a phenotype data frame to join with the growth dataframe

morph_combine <- morph_data %>% group_by(Line, Temperature) %>% 
  summarise(mean_length = mean(mean_major),
            mean_width = mean(mean_minor),
            mean_ar = mean(mean_ar),
            mean_speed = mean(gross_speed))

growth_combine <- growth_data %>% filter(Temperature <= 34)

growth_phenotypes <- left_join(growth_combine, morph_combine, by = c('Line', 'Temperature'))

```

### Length

Now that we have combined the predictions of the morphological and movement data with the growth data across temperatures, we can look and see if there are relationships between plasticity in morphology and movement and growth. We will start by looking at length.

```{r}

ggplot(data = growth_phenotypes, aes(x = mean_length, y = Growth.Rate.Hours)) + geom_point(aes(color = Temperature)) + 
  geom_smooth(method = 'lm', se = FALSE) + facet_wrap(facets = "Line", nrow = 5, ncol = 4) + 
  scale_color_viridis_c()

fit_length_growth <- lm(Growth.Rate.Hours ~ mean_length*Line, data = growth_phenotypes)

summary(fit_length_growth)

fit_length_growth_noint <- lm(Growth.Rate.Hours ~ mean_length + Line, data = growth_phenotypes)

AIC(fit_length_growth, fit_length_growth_noint)

ggplot(data = growth_phenotypes, aes(x = mean_length, y = Growth.Rate.Hours, group = Line)) + geom_point(aes(color = Temperature)) + 
  scale_color_viridis_c() + geom_smooth(method = 'lm', se = FALSE)

### calculate slope of length growth rate relationship for each outcrossed line

length_growth_slope_data <- data.frame(Line = unique(growth_phenotypes$Line),
                              Slope = c(coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG120'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG33'],
                                       coef(fit_length_growth)[2],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG110'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG35'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG41'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG44'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG69'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG16'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG38'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG72'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG19'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG77'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG104'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG89'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG20'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG34'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG30'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG90'],
                                       coef(fit_length_growth)[2] + coef(fit_length_growth)['mean_length:LineG36']))

length_growth_slope_data <- left_join(length_growth_slope_data, summ_length, by = 'Line')


ggplot(data = length_growth_slope_data, aes(x = log(range_length), y = abs(Slope))) + geom_point() + geom_smooth(method = 'lm')

summary(lm(formula = abs(Slope) ~ range_length, data = length_growth_slope_data))


```

In general, we see that there is a negative relationship between length and growth rates across temperatures in each of the genotypes. A statistical model with an interaction between length and and outcrossed line also outperforms a model with no interaction suggesting that there is significant variation among lines in the relationship between body size change and growth rate change. G19 is a bit weird ... 

### Width

```{r}

ggplot(data = growth_phenotypes, aes(x = mean_width, y = Growth.Rate.Hours)) + geom_point(aes(color = Temperature)) + 
  geom_smooth(method = 'lm', se = FALSE) + facet_wrap(facets = "Line", nrow = 5, ncol = 4) + 
  scale_color_viridis_c()

fit_width_growth <- lm(Growth.Rate.Hours ~ mean_width*Line, data = growth_phenotypes)

summary(fit_width_growth)

fit_width_growth_noint <- lm(Growth.Rate.Hours ~ mean_width + Line, data = growth_phenotypes)

AIC(fit_width_growth, fit_width_growth_noint)

```

As for length, we also see that shorter paramecia within the outcrossed lines have higher growth rates. Again, a model with an interaction between width and outcrossed line is preferred over a model without an interaction. 

### Aspect Ratio

```{r}

ggplot(data = growth_phenotypes, aes(x = mean_ar, y = Growth.Rate.Hours)) + geom_point(aes(color = Temperature)) + 
  geom_smooth(method = 'lm', se = FALSE) + facet_wrap(facets = "Line", nrow = 5, ncol = 4) + 
  scale_color_viridis_c()

fit_ar_growth <- lm(Growth.Rate.Hours ~ mean_ar*Line, data = growth_phenotypes)

summary(fit_ar_growth)

fit_ar_growth_noint <- lm(Growth.Rate.Hours ~ mean_ar + Line, data = growth_phenotypes)

AIC(fit_ar_growth, fit_ar_growth_noint)

```

Although maybe not quite as strong as the length or width relationships, it does appear that paramecia within each outcrossed line generally have higher growth rates when they have higher aspect ratios. Again, a model with an interaction between aspect ratio and outcrossed line performs better than a model without an interaction.

### Speed

```{r}

ggplot(data = growth_phenotypes, aes(x = mean_speed, y = Growth.Rate.Hours)) + geom_point(aes(color = Temperature)) + 
  geom_smooth(method = 'lm', se = FALSE) + facet_wrap(facets = "Line", nrow = 5, ncol = 4) + 
  scale_color_viridis_c()

fit_speed_growth <- lm(Growth.Rate.Hours ~ mean_speed*Line, data = growth_phenotypes)

summary(fit_speed_growth)

fit_speed_growth_noint <- lm(Growth.Rate.Hours ~ mean_speed + Line, data = growth_phenotypes)

fit_speed_growth_line <- lm(Growth.Rate.Hours ~ Line, data = growth_phenotypes)

fit_speed_growth_null <- lm(Growth.Rate.Hours ~ 1, data = growth_phenotypes)

AIC(fit_speed_growth, fit_speed_growth_noint, fit_speed_growth_line, fit_speed_growth_null)

```

Although the relationships do not appear as consistent between speed and growth, a linear model with an interaction between speed and outcrossed line is the best performing model even when considering a null model with just an intercept and model that only includes line and does not include speed at all.

### Within line varation and growth

So, what does all of this mean? Well, it appears that within outcrossed lines, changes in size, aspect ratio, and speed are associated with changes in growth rates. In general, smaller, thinner, faster paramecium have higher growth rates. 

## Plasticity in morphology and movement

Next, we want to ask whether the extent to which an outcrossed line is plastic in its morphology and movement is related to the extent to which its growth rate changes across temperatures. First, we will just examine plasticity in each of the phenotypes and whether it is associated with the extent to which growth changed for each outcrossed line. We will quantify plasticity in growth similar to the way that we quantified plasticity in morphology and movement. That is, by averaging the measured growth rates at each of the temperatures and then calculating the range in growth across temperatures. Before, we do that, we will fit a GAM to the TPC data just to examine differences among outcrossed lines.

```{r}

### fit a gam to the growth data

growth_data$Date.paramecium.introduced <- as.factor(growth_data$Date.paramecium.introduced)

tpc_fit <- gam(formula = Growth.Rate.Hours ~ Line + s(Temperature, k = 5, bs = 'tp', by = Line) + s(Date.paramecium.introduced, bs = 're'), data = growth_data, method = 'REML')

summary(tpc_fit)

tpc_fit_noint <- gam(formula = Growth.Rate.Hours ~ Line + s(Temperature, k = 5, bs = 'tp') + s(Date.paramecium.introduced, bs = 're'), data = growth_data, method = 'REML')

AIC(tpc_fit, tpc_fit_noint)

draw(tpc_fit)

```
The GAMs here actually suggest that a model including Line but no interaction between temperature and Line actually performs best with the growth data. This suggests that there is potentially a genetic effect (moving the line up and down), an environmental effect (how the line changes with temperature), but no gene by environment interaction in the TPCs. That being said, we can still get the plasticity of the growth rate for each of the outcrossed lines. 

This apparently changes after accounting for a random effect of day.

```{r}

tpc_means <- growth_data %>% group_by(Line, Temperature) %>% 
  summarise(growth = median(Growth.Rate.Hours)) %>% 
  ungroup() %>% group_by(Line) %>% 
  summarise(mean_growth = mean(growth),
            max_growth = max(growth),
            min_growth = min(growth),
            range_growth = max_growth - min_growth,
            stand_range_growth = range_growth/max_growth)

```

Using the range in growth for the outcrossed lines as our metric of plasticity, we can ask whether the extent of plasticity in any of the morphology or movement variables is predictive of the extent of plasticity in growth rates.

```{r}

plast_growth_data <- full_join(plast_morph_data, tpc_means, by = 'Line')

### length plasticity

ggplot(data = filter(plast_growth_data, Line != "G20"), aes(x = range_length, y = range_growth)) +
  geom_point() + geom_smooth(method = 'lm', se = TRUE)

### width plasticity

ggplot(data = plast_growth_data, aes(x = range_width, y = range_growth)) +
  geom_point() + geom_smooth(method = 'lm', se = TRUE)

ggplot(data = plast_growth_data, aes(x = range_ar, y = range_growth)) +
  geom_point() + geom_smooth(method = 'lm', se = TRUE)

ggplot(data = plast_growth_data, aes(x = range_speed, y = range_growth)) +
  geom_point() + geom_smooth(method = 'lm', se = TRUE)


```
As can be seen from these graphs, there doesn't appear to be any relationship between the extent of phenotypic plasticity in any of the morphological variables and the extent of plasticity in growth. However, as we saw before, the plasticity in size (length/width) is not necessary related to the plasticity in shape and speed. It is possible that we need a measure of multivariate plasticity to compare to the amount of plasticity in growth.

To quantify the amount of multivariate plasticity, we will use a PCA of the morphology data using the means for each line at each temperature. We will then use the area of a convex hull containing those points to be our multivariate measure of plasticity. 

First, we will make perform the PCA.

```{r}

# perform pca using the rda function in vegan

pca2 <- pca(morph_combine[,3:6], scale = TRUE)

summary(pca2)

biplot(pca2)

ordihull(pca2, groups = morph_combine$Line)

hull_data <- t(summary(ordihull(pca2, groups = morph_combine$Line)))

hull_data <- cbind(rownames(hull_data), data.frame(hull_data, row.names=NULL))

colnames(hull_data)[1] <- 'Line'

plast_growth_data <- full_join(plast_growth_data, hull_data, by = 'Line')

ggplot(data = plast_growth_data, aes(x = Area, y = range_growth)) + geom_point() + geom_smooth(method = 'lm')

summary(lm(range_growth ~ Area, data = plast_growth_data))

```

I think our original hypothesis here was that morphological and movement plasticity would potentially allow the paramecia to buffer the effects of temperature and that this would lead to less plasticity in growth with temperature. The data, however, seem to show that there is no relationship between the degree of change in morphology and movement and the degree of change in growth rates. This feels a bit strange given that we do see relationships between things like length and aspect ratio and growth. I am currently thinking that this lack of relationship means that within an outcrossed line, size changes (for example) are associated with changes in growth rates. But, across outcrossed lines, some lines change their size a little whereas some lines change their sizes a lot and yet those two lines may show the same degree of growth rate change.

## Conclusions

To wrap all of this up, what have we done and found?

In response to changes in temperature, the paramecia do exhibit phenotypic plasticity. In general, paramecium get shorter, less wide, skinnier and faster as temperatures increase, but this relationship is often nonlinear with evidence for a switch in this pattern at the highest temperatures. Moreover, changes in length and width and strongly correlated with one another as are changes in aspect ratio and speed. The correlations in traits between these groups is less strong which is also true for genetic variation among outcrossed lines. We also find that these morphological and movement changes are associated with changes in paramecium growth within outcrossed lines. That is, as the paramecia within outcrossed lines get smaller, skinnier, and faster, they also tend to have higher growth rates. However, across outcrossed lines, we do not find a relationship between the extent to which morphology and movement change and the total extent to which paramecium growth rates change either from a univariate or multivariate view of their morphological and movement changes. Looking forward to folks' thoughts about this.















